---
title: "Data Analysis Replication"
author: "Prabhani Athukorala"
format: html
editor: visual
---

## Introduction

This report presents a replication of selected data analyses from the study "**Auditory and cognitive contributions to recognition of degraded speech in noise: Individual differences among older adults"** by **Daniel Fogerty and Judy R. Dubno.**

This study belongs to the field of auditory neuroscience and speech perception.

The goal of this replication is to reproduce a subset of key statistical findings in the above study using R and to evaluate how closely the replicated results align with those reported in the original study.

## Original Study Overview

#### 1) Goal of the study

-   This study looked at how older adults' auditory and cognitive capacities affect their ability to recognize speech in noisy environments.

-   The study specifically examined individual differences in speech recognition skills between older people with hearing impairment (OHI) and those with. normal hearing (ONH).

-   To understand degraded speech under difficult listening situations, the authors attempted to identify which underlying skills such as hearing sensitivity, working memory, vocabulary knowledge, and speech "glimpsing" ability are most crucial.

#### 2) Data used in the original study

Participants: 40 adults (20 with normal hearing (ONH) and 20 with hearing impairment (OHI))

Predictor variables:

| **Auditory measures** | Auditory and Cognitive measures |
|------------------------------------|------------------------------------|
| Hearing thresholds | Tasks assessing working memory, episodic memory, processing speed, and executive function |
| Speech related processing abilities (modulation detection and speech glimpsing) | Cognitive linguistic measures such as vocabulary knowledge and linguistic closure |

Dataset

Three experimental degraded speech conditions were considered.

1.  Spectrally reduced speech

2.  Speech in modulated noise

3.  A combination of both

For each data set above, speech recognition thresholds (SRTs) were calculated at 20%, 50%, and 80% recognition levels. The three datasets described in the original study do not refer to separate data files, but rather to three different experimental speech recognition conditions. These are represented in the dataset by separate outcome variables (D1, D2, D3), each corresponding to a different type of degraded speech condition.

The method used to do that was psychometric functions based on the extended Short-Time Objective Intelligibility (eSTOI). ( This is a mathematical mapping model that relate the objective eSTOI score (a scalar between 0 and 1) to subjective human speech intelligibility scores).

#### 3) Analyses conducted in the original study

Like I mentioned above, the original study looked at the association between predictor variables and speech recognition performance. the statistical methods they used are:

1\) A Principal components analysis (PCA) to generate composite predictor variables and minimize the dimensionality of auditory and cognitive assessments. (The data set after PCA is available for use ready)

2\) After PCA, a summarization of participant characteristics and key predictor variables for the ONH and OHI groups have been done.

2\) A correlational analyses to see which predictors were linked to speech recognition results. A subset of important predictors was chosen for additional examination based on these findings.

3\) A multiple regression dominance analysis to get certain the relative significance of each predictor in explaining variance in speech recognition performance.

-   This analysis was carried out independently for the ONH and OHI groups, as well as for various datasets and task difficulty levels (SRT20, SRT50, and SRT80).

#### 4) Analyses Chosen for Replication

I choose to replicate the following analyses from above except for the initial exploratory data analysis i am going to do.

1.  First, descriptive statistics are used to summarize participant characteristics and key predictor variables for the ONH and OHI groups. I do this analysis using variables of age, gender, hearing sensitivity.

2.  Second, I will create a descriptive visualization is created using violin plots to compare the distributions of key predictor variables between groups, similar to Figure 3 in the original study.

3.  Third, I will conduct an inferential analysis to examine the relationship between predictor variables and speech recognition outcomes, providing a simplified approximation of the dominance analysis used in the original study.I will do a visualization of this analysis too.

4.  Finally, a I will do a correlation analysis is performed to explore relationships between auditory and cognitive predictors and speech recognition performance, and to assess whether similar patterns to those reported in the original study are observed. This will be a second descriptive analysis.

## Variables used in the study

## Data import and initial inspection, visualization of data

The dataset I am going to use for this replication, was obtained from the supporitng information provided with the original paper.

The data file contains participant level measurements auditory and cognitive predictor variables, and speech recognition outcomes.

1\) I will initially inspect the structure of the dataset and the relationships among variables. What I want to show is how these datasets contain large number of variables and why they expected redundancy among them and why it was necessary to first reduce the dimensionality of the data.

```{r}

# Installing and loading the necessary libraries

install.packages("corrplot")

library(readxl)
library(tidyverse)
library(corrplot)
```

First I am going to load the data sheets for auditory and cognitive measures in **Table 1. Summary table of experimental measures, including domains, tasks, and variables.** <https://doi.org/10.1371/journal.pone.0331487.t001>

```{r}
auditory_variables <- read_excel("data/journal.pone.0331487.s002.xlsx",sheet = "AuditoryVariables",skip = 1) # Skipping the header above data columns because the auditory dataset contained multi-level headers

cognitive_variables <- read_excel("data/journal.pone.0331487.s002.xlsx",
                        sheet = "CognitiveVariables (z-scores)")
```

First, doing a basic exploration

```{r}
head(auditory_variables)
glimpse(auditory_variables)
```

```{r}
head(cognitive_variables)
glimpse(cognitive_variables)
```

Now I am going to draw some distribution plots for the exploratory data analysis. There are no plots shown for this in the original paper. So I am going to draw.

For auditory variables,

```{r}
auditory_variables_long <- auditory_variables%>%
  pivot_longer(cols = `250 Hz`:`8000 Hz`,
               names_to = "Frequency",
               values_to = "Threshold")

ggplot(auditory_variables_long, aes(x = Threshold)) +
  geom_histogram(bins = 20) +
  facet_wrap(~Frequency, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Hearing Thresholds Across Frequencies")
```

Correlations between auditory variables

```{r}
auditory_variables_1 <- auditory_variables %>%
  select(`250 Hz`:`8000 Hz`)

auditory_correlations <- cor(auditory_variables_1)

library(corrplot)
corrplot(auditory_correlations, method = "color", type = "upper")
```

For cognitive variables

```{r}
cognitive_variables_long <- cognitive_variables %>%
  pivot_longer(cols = zShortRecall:zStroop_Mismatch,
               names_to = "Test",
               values_to = "Score")

ggplot(cognitive_variables_long, aes(x = Score)) +
  geom_histogram(bins = 20) +
  facet_wrap(~Test, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Cognitive Test Scores")
```

Correlations between cognitive variables

```{r}
cognitive_variables_1 <- cognitive_variables %>%
  select(zShortRecall:zStroop_Mismatch)

cognitive_correlation <- cor(cognitive_variables_1)

# drawing the correlation matrix

corrplot(cognitive_correlation, method = "color", type = "upper")
```

The correlation matrices shows a substantial correlations among variables within both auditory and cognitive domains. In the auditory dataset, hearing thresholds across different frequencies show strong positive correlations, indicating that individuals with elevated thresholds at one frequency tend to show similar patterns across other frequencies. This suggests redundancy in the measurement of hearing sensitivity.

Similarly, in the cognitive dataset, several measures show moderate to strong correlations, particularly among tasks assessing related cognitive functions such as memory and executive control. These patterns indicate that multiple variables may be capturing overlapping underlying constructs.

These findings support the use of principal components analysis (PCA) in the original study, as PCA allows for the reduction of correlated variables into a smaller set of independent components representing underlying auditory and cognitive processes.

The dataset includes principal component (PCA) scores for auditory and cognitive predictors, as well as speech recognition thresholds (SRTs) for multiple datasets. Because PCA variables are already provided, I am not going to recompute PCA, and instead I use these variables directly in the analyses.

## Replication Analysis 1: Descriptive Statistics

### Descriptive Analysis: Participant characteristics

The first descriptive analysis in the original paper focuses on describing the demographic and baseline characteristics of the participants. This includes summarizing age, gender distribution, hearing thresholds, for the two groups: older normal-hearing (ONH) and older hearing-impaired (OHI) listeners.

The goal of this analysis is to replicate the general patterns reported in the original study, such as differences in hearing thresholds between groups and similarities in cognitive abilities.

Now I am loading the correct sheet where the data after doing the PCA., which is named as "Predictors&Outcomes"/\].

```{r}
df <- read_excel(
  "data/journal.pone.0331487.s002.xlsx",
  sheet = "Predictors&Outcomes", ,skip = 1
)
df$HearGroup <- as.factor(df$HearGroup)

```

Summarizing age and sample size

```{r}
df %>%
  group_by(HearGroup) %>%
  summarise(
    n = n(),
    Age_mean = mean(Age),
    Age_sd = sd(Age),
    Age_min = min(Age),
    Age_max = max(Age)
  )
```

Gender distribution

```{r}
table(df$HearGroup, df$Sex)
```

Key predictor variables : Audiogram_PCA ; Glimpsing ; WorkingMemory_PCA; Vocabulary. The authors have not generated any kind of summarization or any descriptive statistics out of these variables. But, they have analyzed a frequency by frequency mean audiogram from raw audiometric threshold columns like 250Hz, 500Hz, 1000Hz and so on from the "AuditoryVariables" sheet.

So I am trying to replicate that.

for that I need to seperate a dataset that only ha sthe threshold columns.

```{r}

auditory_variables_long <- auditory_variables %>%
  select(HearGroup, `250 Hz`, `500 Hz`, `1000 Hz`, `2000 Hz`, `4000 Hz`, `8000 Hz`) %>%
  pivot_longer(
    cols = `250 Hz`:`8000 Hz`,
    names_to = "Frequency",
    values_to = "Threshold"
  ) %>%
  mutate(
    Frequency = factor(
      Frequency,
      levels = c("250 Hz", "500 Hz", "1000 Hz", "2000 Hz", "4000 Hz", "8000 Hz")
    )
  )

auditory_variables_summary <- auditory_variables_long %>%
  group_by(HearGroup, Frequency) %>%
  summarise(
    mean_threshold = mean(Threshold, na.rm = TRUE),
    sd_threshold = sd(Threshold, na.rm = TRUE),
    .groups = "drop"
  )

auditory_variables_summary
```

Now I can make a scatter kind of plot also using the sd_threshold just like in the original study.

The speciality is when plotting an audiogram we have to make the y axis reverse, because lower thresholds are the better and higher values.

```{r}
ggplot(aud_summary, aes(x = Frequency, y = mean_threshold,
                        group = HearGroup, shape = HearGroup, linetype = HearGroup)) +
  geom_line(color = "black") +
  geom_point(color = "black", size = 2.5) +
  geom_errorbar(
    aes(ymin = mean_threshold - sd_threshold, ymax = mean_threshold + sd_threshold),
    width = 0.1,
    color = "black"
  ) +
  scale_y_reverse() +
  labs(
    title = "Mean Audiograms for the Two Listener Groups",
    x = "Frequency (Hz)",
    y = "Threshold (dB HL)"
  ) +
  theme_classic()
```

### Replication Analysis 2: Descriptive Visualization

To visualize a part of the above descriptive statistics, I will try to create the violin plots to compare the distributions of key predictor variables between ONH and OHI groups. The selected variables include:

-   Audiogram PCA (hearing sensitivity) - now not the raw threshold values but value s after PCA
-   Speech glimpsing ability
-   Working memory PCA
-   Vocabulary knowledge

These all values are extracted from the same after PCA data sheet.

To create the violin plots,

```{r}

df$HearGroup <- factor(df$HearGroup, levels = c("ONH", "OHI"))

plot_df <- df %>%
  select(HearGroup, Audiogram_PCA, Glimpsing, WorkingMemory_PCA, Vocabulary) %>%
  pivot_longer(
    cols = c(Audiogram_PCA, Glimpsing, WorkingMemory_PCA, Vocabulary),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = factor(
      Variable,
      levels = c("Audiogram_PCA", "Glimpsing", "WorkingMemory_PCA", "Vocabulary"),
      labels = c("Audiogram PCA", "Glimpsing (z-score)", "Working Memory PCA", "Vocab.Knowledge -z-score")
    )
  )

ggplot(plot_df, aes(x = HearGroup, y = Value)) +
  geom_violin(trim = FALSE, fill = "grey80", color = "black") +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  stat_summary(fun = median, geom = "point", size = 2) +
  facet_wrap(~Variable, scales = "free_y", nrow = 1) +
  labs(
    x = NULL,
    y = NULL,
    title = "Violin plots for the four predictor variables"
  ) +
  theme_classic()
```

### Replication Analysis 3: Inferential Analysis: Predictor Importance

Now I am going to replicate the relative contribution of predictor variables to speech recognition performance. The original study used dominance analysis to quantify the proportion of variance explained by each predictor across datasets and performance levels.

In this replication, a simplified inferential approach will be used to approximate these relationships by examining how key predictors (Audiogram PCA, glimpsing, working memory, and vocabulary) relate to speech recognition outcomes. This allows for evaluation of whether similar patterns emerge, such as the stronger role of vocabulary for ONH listeners and glimpsing for OHI listeners.

### Replication Analysis 4: Inferential Visualization

To support interpretation of predictor importance, a visualization (e.g., a bubble chart or comparable summary plot) will be used to represent the relative contributions of different predictors. This is conceptually similar to the dominance analysis visualizations presented in the original study.

### Replication Analysis 5: Second Descriptive Analysis: Correlation Analysis

Finally, a correlation analysis will be conducted to examine relationships between predictor variables and speech recognition outcomes. This analysis aims to replicate patterns reported in the original study, where:

-   Audiogram PCA, speech glimpsing, working memory, and vocabulary were consistently associated with speech recognition performance
-   Audiogram PCA and glimpsing showed stronger relationships in OHI listeners
-   Vocabulary showed stronger relationships in ONH listeners

This analysis provides additional insight into how auditory and cognitive factors interact in predicting speech recognition and further supports the interpretation of the inferential analyses.

## Comparison to the Original Paper - how successful were you at replicating the analyses and visualizations in the study. What problems did you encounter? Why might you have encountered those problems? What details were lacking from the original study's methods that might have hampered your ability to replicate the authors' results?\]

## Discussion and Reflection

## References
