---
title: "Data Analysis Replication"
author: "Prabhani Athukorala"
format: html
editor: visual
---

## Introduction

This report presents a replication of selected data analyses from the study "**Auditory and cognitive contributions to recognition of degraded speech in noise: Individual differences among older adults"** by **Daniel Fogerty and Judy R. Dubno.**

This study belongs to the field of auditory neuroscience and speech perception.

The goal of this replication is to reproduce a subset of key statistical findings in the above study using R and to evaluate how closely the replicated results align with those reported in the original study.

## Original Study Overview

#### 1) Goal of the study

-   This study looked at how older adults' auditory and cognitive capacities affect their ability to recognize speech in noisy environments.

-   The study specifically examined individual differences in speech recognition skills between older people with hearing impairment (OHI) and those with. normal hearing (ONH).

-   To understand degraded speech under difficult listening situations, the authors attempted to identify which underlying skills such as hearing sensitivity, working memory, vocabulary knowledge, and speech "glimpsing" ability are most crucial.

#### 2) Data used in the original study

Participants: 40 adults (20 with normal hearing (ONH) and 20 with hearing impairment (OHI))

Predictor variables:

| **Auditory measures** | Auditory and Cognitive measures |
|----|----|
| Hearing thresholds | Tasks assessing working memory, episodic memory, processing speed, and executive function |
| Speech related processing abilities (modulation detection and speech glimpsing) | Cognitive linguistic measures such as vocabulary knowledge and linguistic closure |

Data sets: Three experimental data sets with three degraded speech conditions

1.  Spectrally reduced speech

2.  Speech in modulated noise

3.  A combination of both

For each data set above, speech recognition thresholds (SRTs) were calculated at 20%, 50%, and 80% recognition levels. The method used to do that was psychometric functions based on the extended Short-Time Objective Intelligibility (eSTOI). ( This is a mathematical mapping model that relate the objective eSTOI score (a scalar between 0 and 1) to subjective human speech intelligibility scores).

#### 3) Analyses conducted in the original study

Like I mentioned above, the original study looked at the association between predictor variables and speech recognition performance. the statistical methods they used are:

1\) A Principal components analysis (PCA) to generate composite predictor variables and minimize the dimensionality of auditory and cognitive assessments. (The data set after PCA is available for use ready)

2\) A correlational analyses to see which predictors were linked to speech recognition results. A subset of important predictors was chosen for additional examination based on these findings.

3\) A multiple regression dominance analysis to get certain the relative significance of each predictor in explaining variance in speech recognition performance.

-   This analysis was carried out independently for the ONH and OHI groups, as well as for various datasets and task difficulty levels (SRT20, SRT50, and SRT80).

#### 4) Analyses Chosen for Replication

I choose to replicate the following analyses from above except for the initial exploratory data analysis i am going to do.

-   Descriptive statistics - to summarize key variables for each group (ONH and OHI), including demographic variables and selected auditory and cognitive predictors.

-   Inferential analysis - to examine group differences between ONH and OHI participants on selected predictor variables using independent samples t-tests.

-   A visualization - to explore relationships between predictor variables and speech recognition outcomes, using a correlation based plot.

-   A second inferential analysis - using multiple linear regression to examine how selected auditory and cognitive predictors contribute to speech recognition performance (SRT50).

## Variables used in the study

## Data import and initial inspection, visualization of data

The dataset I am going to use for this replication, was obtained from the supporitng information provided with the original paper.

The data file contains participant level measurements auditory and cognitive predictor variables, and speech recognition outcomes.

1\) I will initially inspect the structure of the dataset and the relationships among variables. What I want to show is how these datasets contain large number of variables and why they expected redundancy among them and why it was necessary to first reduce the dimensionality of the data.

```{r}

# Installing and loading the necessary libraries

install.packages("corrplot")

library(readxl)
library(tidyverse)
library(corrplot)
```

First I am going to load the data sheets for auditory and cognitive measures in **Table 1. Summary table of experimental measures, including domains, tasks, and variables.** <https://doi.org/10.1371/journal.pone.0331487.t001>

```{r}
auditory_variables <- read_excel("data/journal.pone.0331487.s002.xlsx",sheet = "AuditoryVariables",skip = 1) # Skipping the header above data columns because the auditory dataset contained multi-level headers

cognitive_variables <- read_excel("data/journal.pone.0331487.s002.xlsx",
                        sheet = "CognitiveVariables (z-scores)")
```

First, doing a basic exploration

```{r}
head(auditory_variables)
glimpse(auditory_variables)
```

```{r}
head(cognitive_variables)
glimpse(cognitive_variables)
```

Now I am going to draw some distribution plots for the exploratory data analysis. There are no plots shown for this in the original paper. So I am going to draw.

For auditory variables,

```{r}
auditory_variables_long <- auditory_variables%>%
  pivot_longer(cols = `250 Hz`:`8000 Hz`,
               names_to = "Frequency",
               values_to = "Threshold")

ggplot(auditory_variables_long, aes(x = Threshold)) +
  geom_histogram(bins = 20) +
  facet_wrap(~Frequency, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Hearing Thresholds Across Frequencies")
```

Correlations between auditory variables

```{r}
auditory_variables_1 <- auditory_variables %>%
  select(`250 Hz`:`8000 Hz`)

auditory_correlations <- cor(auditory_variables_1)

library(corrplot)
corrplot(auditory_correlations, method = "color", type = "upper")
```

For cognitive variables

```{r}
cognitive_variables_long <- cognitive_variables %>%
  pivot_longer(cols = zShortRecall:zStroop_Mismatch,
               names_to = "Test",
               values_to = "Score")

ggplot(cognitive_variables_long, aes(x = Score)) +
  geom_histogram(bins = 20) +
  facet_wrap(~Test, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Cognitive Test Scores")
```

Correlations between cognitive variables

```{r}
cognitive_variables_1 <- cognitive_variables %>%
  select(zShortRecall:zStroop_Mismatch)

cognitive_correlation <- cor(cognitive_variables_1)

# drawing the correlation matrix

corrplot(cognitive_correlation, method = "color", type = "upper")
```

The correlation matrices shows a substantial correlations among variables within both auditory and cognitive domains. In the auditory dataset, hearing thresholds across different frequencies show strong positive correlations, indicating that individuals with elevated thresholds at one frequency tend to show similar patterns across other frequencies. This suggests redundancy in the measurement of hearing sensitivity.

Similarly, in the cognitive dataset, several measures show moderate to strong correlations, particularly among tasks assessing related cognitive functions such as memory and executive control. These patterns indicate that multiple variables may be capturing overlapping underlying constructs.

These findings support the use of principal components analysis (PCA) in the original study, as PCA allows for the reduction of correlated variables into a smaller set of independent components representing underlying auditory and cognitive processes.

The dataset includes principal component (PCA) scores for auditory and cognitive predictors, as well as speech recognition thresholds (SRTs) for multiple datasets. Because PCA variables are already provided, I am not going to recompute PCA, and instead I use these variables directly in the analyses.

## Replication Analysis 1: Descriptive Statistics

## Replication Analysis 2: Inferential Analysis

## Replication Analysis 3: Visualization

## Replication Analysis 4: Second Inferential Analysis

## Comparison to the Original Paper

## Discussion and Reflection

## References
